---
title: 'NBA box score analysis'
output: html_document
author: "Serdar Altok"
date: "`r format(Sys.Date(), '%m/%d/%y')`"
---

```{r set options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css styling, echo=FALSE}

<style>
.tocify {
max-width: 175px;
}
</style>

<style>
.main-container {
width: 100%;
max-width: 940px;
margin-left: 250px;
margin-right: auto;
}
</style>

<style>
.red-header {
  color: red;
}
</style>

```



# Introduction

In this project, I will explore some stats of interest derived from nba box score data and build a model for playoff success. My goal is to identify factors that can be improved upon team building and player development, rather than guess the number of wins. 

## Setup and Data

```{r load data, message = F, warning = F}
library(tidyverse)

player_data <- read_csv("player_game_data.csv")
team_data <- read_csv("team_game_data.csv")
```

```{r}

# split team data to regular season and playoff
team_reg <- team_data %>% filter(gametype == 2)
team_playoff <- team_data %>% filter(gametype == 4)

```

## Calculate advanced stats

```{r}

# Formulas for advanced stats

# effective field goal percentage (eFG)
calc_efg <- function(fgm, fgm3, fga) {
  efg <- (fgm + 0.5 * fgm3) / fga * 100
  return(round(efg, 1))
}

# points per attempt (ppa)
calc_ppa <- function(shotattemptpoints, shotattempts) {
  return(round(shotattemptpoints/shotattempts, 1))
}

# asssist percentage (assist_pct)
calc_assist_pct <- function(assists, fgm) {
  return(round(100*(assists / fgm), 1))
}

# turnover percentage (tov_pct)
calc_tov_pct <- function(turnovers, shotsattempts) {
  return(round(100*(turnovers / (turnovers + shotsattempts)), 1))
}

# defensive rebound percentage (dreb_pct)
calc_dreb_pct <- function(def_rebdefensive, def_reboundchance) {
  return(round(100*(def_rebdefensive / def_reboundchance), 1))
}

# offensive rebound percentage (oreb_pct)
calc_oreb_pct <- function(reboffensive, reboundchance) {
  return(round(100*(reboffensive / reboundchance), 1))
}

# steal percentage (steal_pct)
calc_steal_pct <- function(steals, def_possessions) {
  return(round(100*(steals / def_possessions), 1))
}

# block percentage (block_pct)
calc_block_pct <- function(blocks, def_fg2attempted) {
  return(round(100*(blocks / def_fg2attempted), 1))
}

```

```{r}
# Add defensive stats

info_cols <- c('nbagameid', 'off_team', 'def_team')
def_stats <- c('points', 'fg2attempted', 'possessions', 'rebdefensive', 'reboundchance', 'reboffensive', 'stealsagainst', 'blocksagainst', 'fgmade', 'fg3made', 'fgattempted')

team_reg_defensive <- team_reg %>%
  select(all_of(c(info_cols, def_stats))) %>%
  rename(def_points = points,
         def_fgmade = fgmade, 
         def_fg3made = fg3made,
         def_fgattempted = fgattempted,
         def_fg2attempted = fg2attempted,
         def_possessions = possessions,
         def_reboffensive = reboffensive,
         def_rebdefensive = rebdefensive,
         def_reboundchance = reboundchance,
         steals = stealsagainst,
         blocks = blocksagainst
         )

# after joining the defensive stats, the def_ prefix indicates that these are the defensive stats of the offensive team, i.e. when they are on defense

team_reg <- team_reg %>% 
  inner_join(team_reg_defensive, 
             by=c('nbagameid'='nbagameid', 
                  'off_team'='def_team', 
                  'def_team'='off_team'))
```

```{r}

# Add eFG and def_eFG stats

team_reg <- team_reg %>%
  mutate(
    eFG = calc_efg(fgmade, fg3made, fgattempted),
    def_eFG = calc_efg(def_fgmade, def_fg3made, def_fgattempted))

# Add differential stats
team_reg <- team_reg %>% 
  mutate(diff_points = points - def_points,
         diff_reboffensive = reboffensive - def_reboffensive,
         diff_possessions = possessions - def_possessions,
         diff_eFG = eFG - def_eFG
         )
```

```{r}

# keep relevant columns and calculate totals

stats_summed <- c('shotattemptpoints', 'shotattempts', 
                  'points', 'def_points', 'possessions', 'def_possessions',
                  'fgmade','fg3made','fgattempted',
                  'def_fgmade', 'def_fg3made', 'def_fgattempted',
                  'assists' , 'steals', 'blocks', 'turnovers',
                  'def_possessions', 'def_fg2attempted',
                  'reboffensive', 'reboundchance', 
                  'def_rebdefensive', 'def_reboundchance')
diff_stats <- c( "diff_possessions", "diff_eFG")

# calculate offensive and defensive totals, means
team_reg_totals <- team_reg %>%
  group_by(season, off_team) %>%
  summarize(
    across(all_of(stats_summed), sum),
    across(all_of(c('off_win', diff_stats)), mean),
    .groups = 'drop'
  )

# Calculate advanced stats

team_reg_totals <- team_reg_totals %>%
  mutate(
    win_pct = off_win,
    eFG = calc_efg(fgmade, fg3made, fgattempted),
    def_eFG = calc_efg(def_fgmade, def_fg3made, def_fgattempted),
    ppa = calc_ppa(shotattemptpoints, shotattempts),
    assist_pct = calc_assist_pct(assists , fgmade),
    steal_pct = calc_steal_pct(steals, def_possessions) ,
    block_pct = calc_block_pct(blocks , def_fg2attempted),
    tov_pct = calc_tov_pct(turnovers, shotattempts),
    oreb_pct = calc_oreb_pct(reboffensive, reboundchance),
    dreb_pct = calc_dreb_pct(def_rebdefensive, def_reboundchance)
  )

```

```{r}

# Offensive eFG for Golden State Warriors
print(paste('GSW offensive eFG% in 2015-2016 regular season is', 
            team_reg_totals %>%
              filter(season == 2015, off_team == 'GSW') %>%
              pull(eFG)))

# Defensive eFG for Golden State Warriors
print(paste('GSW defensive eFG% in 2015-2016 regular season is', 
            team_reg_totals %>%
              filter(season == 2015, off_team == 'GSW') %>%
              pull(def_eFG)))

```

## Win frequency of the team with higher EFG%

```{r}
# each game is counted twice, won't effect percentages

# team with better eFG wins if the signs of points_diff and eFG_diff are same
higher_eff_wins_pct <- team_reg %>%
  filter(diff_eFG != 0) %>%
  mutate(p = (diff_points > 0) == (diff_eFG > 0)) %>%
  summarise(win_pct = mean(p)) %>%
  pull(win_pct)

# Print the result rounded to 1 decimal place
print(round(100 * higher_eff_wins_pct, 1))
```

## Win frequency of the team with more offensive rebounds

```{r}

# Calculate the percentage where the signs of diff_points and diff_reboffensive are the same
higher_reboff_win_pct <-  team_reg %>%
  filter(diff_reboffensive != 0) %>%
  mutate(p = (diff_points > 0) == (diff_reboffensive > 0)) %>%
  summarise(win_pct = mean(p)) %>%
  pull(win_pct)

# Print the result rounded to 1 decimal place
print(round(100*higher_reboff_win_pct, 1))

```


Number of offensive rebounds do not carry any direct information about how well a team is scoring. Some teams missing more shots (a poor offensive performance) migtht be getting more opportunities for an offensive rebound. Or some teams may be chasing offensive rebounds at the cost of conceding more points in transition.

On the other hand, eFG% tells us which team is scoring at a higher rate, which is directly related to the final score. It is also a function of one team's offense and the other team's defense, giving info about both teams' main functions. It doesn't measure free throws or number of extra possessions of one team but is much more informative than the number of offensive rebounds.

## Percentage of games players are available

```{r}
# Add flags for whether player played in the game and whether he is available
player_data <- player_data %>%
  mutate(played = seconds > 0,
         available = 1 - missed)

# Split player data into regular season and playoff games
player_reg <- player_data %>%
  filter(gametype == 2)

player_playoff <- player_data %>%
  filter(gametype == 4)

# Calculate regular season totals for games played, points, and percentage of available games
player_availablity <- player_reg %>%
  group_by(season, player_name) %>%
  summarise(available_pct = mean(available),
            total_points = sum(points),
            total_played = sum(played)) %>%
  ungroup() %>%
  mutate(ppg = total_points / total_played)

# Apply the constraints
available_pct <- player_availablity %>%
  filter(ppg >= 25, available_pct >= 0.25) %>%
  summarise(mean_available_pct = mean(available_pct),
            .groups = 'drop') %>%
  pull(mean_available_pct)

# Calculate the mean available game percentage per season
print(paste0("Mean available game percentage per season: ", 
            round(100 * available_pct, 1),
            "%"))

```

**Notes:**

-   I took the average of seasonal availability percentages. Since some seasons have different number of games, the result would be slightly different if I calculated total number of games available/total number of games.
-   I used games in the player game data to calculate availability percentages. For two players (Kyrie Irving and Bradley Beal 2019 season) the number of listed games is less than the number of games their teams played. If the number of games the team played is used, the result would be slightly different.

## home court team winning percentage in playoffs by round

```{r}
# split playoff data
team_playoff <- team_data %>% filter(gametype == 4)

# Sort by gamedate to identify first and last games of a playoff series
team_playoff <- team_playoff %>% 
  arrange(gamedate)

# Keep one row for each game, and keep relevant columns
team_playoff <- team_playoff %>% 
  distinct(nbagameid, .keep_all = TRUE) %>% 
  select(season, nbagameid, gamedate, off_team, def_team, off_home, off_win)

# Create a new column for playoff series name containing team names and year
team_playoff <- team_playoff %>% 
  rowwise() %>% 
  mutate(series_name = paste0(sort(c(off_team, def_team)), collapse = "") %>% paste0(season)) %>% 
  ungroup()

# Create a column for the winning team name for each game
team_playoff <- team_playoff %>% 
  mutate(winner = ifelse(off_win == 1, off_team, def_team))

# Use first games to identify home court advantage and rounds
first_games <- team_playoff %>% 
  distinct(series_name, .keep_all = TRUE)

# The first 8 games of each playoff are Round 1, next 4 are Round 2, next 2 are Conference Finals, next one is Finals
rounds_season <- c(rep("Round 1", 8), rep("Round 2", 4), rep("Conference Finals", 2), "Finals")
first_games$Round <- rep(rounds_season, 19)

# Home side on the first game has home court advantage
first_games <- first_games %>% 
  mutate(home_court_advantage = ifelse(off_home == 1, off_team, def_team))

# Summarize info
playoff_summary <- first_games %>%
  select(series_name, season, off_team, def_team, Round, home_court_advantage)

# Winner of the last game is the series winner
series_winners <- team_playoff %>% 
  group_by(series_name) %>% 
  slice_tail(n = 1) %>%  # Select the last row of each group
  ungroup() %>% 
  select(series_name, series_winner = winner)

# Add series winners to the summary
playoff_summary <- playoff_summary %>% 
  left_join(series_winners, by = "series_name")

# Check whether home team wins
playoff_summary <- playoff_summary %>% 
  mutate(home_court_ad_wins = home_court_advantage == series_winner)

# Report the average number of wins per round between 2014-2023
playoff_summary %>%
  filter(season %in% 2014:2023) %>%
  group_by(Round) %>%
  summarise(win_pct = round(mean(home_court_ad_wins) * 100, 1))

```



## Percentage of teams with +5 or more net rating that made the second round in the playoffs the following season

### Teams with +5 or more net rating

```{r}

# calculate offensive, defensive and net ratings

calc_rtg <- function(points, possessions) {
  return(round(100 * points / possessions, 2))
}

team_reg_totals <- team_reg_totals %>% 
  mutate(ORTG = calc_rtg(points, possessions),
         DRTG = calc_rtg(def_points, def_possessions),
         netRTG = ORTG - DRTG) %>%
  ungroup()

# take the teams with net rating >= 5
teams_plus5_net_rtg <- team_reg_totals %>%
  filter(netRTG >= 5)

```

### Teams that make 2nd round

```{r}
# Split playoff summary to two, one for off teams, one for def teams

playoff_off_teams <- playoff_summary[c('season', 'off_team', 'Round', 'series_name')]

playoff_def_teams <- playoff_summary[c('season', 'def_team', 'Round', 'series_name')] %>%
  rename(off_team=def_team)

# Combine the two and filter for Round 2
round2_teams <- rbind(playoff_off_teams, playoff_def_teams) %>%
  rename(playoff_season=season) %>%
  filter(Round=='Round 2')

# Define playoff season in teams_plus5_net_rtg
teams_plus5_net_rtg <- teams_plus5_net_rtg %>% 
  mutate(playoff_season = season + 1)

# Find the teams that are +5 net rating and in 2nd round the following year
round2_teams_plus5_net_rtg <- inner_join(round2_teams,
                                         teams_plus5_net_rtg,
                                         by = c("playoff_season", "off_team"))

# Calculate the percentage
second_round_pct <-
  round(100 * nrow(round2_teams_plus5_net_rtg) / nrow(teams_plus5_net_rtg),
        1)

print(
  paste(
    "Percentage of teams with +5 or more net rating that made the second round",
    "in the playoffs the following season is",
    second_round_pct
  )
)

```

## Percentage of players with top five minutes who played in round 2 of playoffs next year

### top 5 players from each team with most minutes in regular season

```{r}

# Get minute data in regular season games
player_min_reg <- player_data %>%
  filter(gametype == 2 & season %in% 2014:2021) %>%
  select(season, team, player_name, seconds)

# Calculate total for time played
player_min_totals <- player_min_reg %>%
  group_by(season, team, player_name) %>%
  summarise(total_secs = sum(seconds), .groups='drop')

# Get top 5 players for each season team combination
player_min_top_5 <- player_min_totals %>%
  arrange(desc(total_secs)) %>%
  group_by(season, team, .groups='drop') %>%
  slice_head(n = 5) %>%
  ungroup()

# Create column to merge with playoff data
player_min_top_5 <- player_min_top_5 %>%
  mutate(playoff_season = season + 1)

# Merge with playoff data for the following year
player_min_top_5 <- player_min_top_5 %>%
  inner_join(round2_teams_plus5_net_rtg, by = c("playoff_season", "team" = "off_team"))
  
```

### find the players who played in round 2

```{r}

# Create a new column for playoff series name containing team names and year
player_playoff <- player_playoff %>%
  rowwise() %>%
  mutate(series_name = paste0(sort(c(team, opp_team)), collapse = "") %>% paste0(season)) %>%
  ungroup()
 
# Find players who played in the playoffs, with total seconds > 0
qualified_players <- player_playoff %>%
  group_by(series_name, player_name, team) %>%
  summarise(total_time = sum(seconds), .groups='drop') %>%
  filter(total_time > 0) %>%
  inner_join(player_min_top_5, by = c("series_name", "player_name", "team")) %>%
  arrange(series_name)

qualified_players_pct <- round(100 * nrow(qualified_players) / nrow(player_min_top_5), 1)

print(
  paste(
    "Percentage of players with top five minutes who played in round 2 of playoffs next year is",
    qualified_players_pct
  )
)

```


## Playoffs Series Modeling


```{r}
library(glmnet)

model_vars = c("eFG", "def_eFG", 'assist_pct', 
                "steal_pct",  "tov_pct", "oreb_pct", "dreb_pct", 'block_pct')

target <- 'win_pct'
team_info <- c('season', 'off_team')

# add reg season totals to model and take competitive teams
model_data <- team_reg_totals %>%
  select(all_of(c(team_info, model_vars, target))) %>%
  filter(win_pct >= .40, season %in% 2014:2022)

X <- as.matrix(model_data[, model_vars])
y <- model_data$win_pct

lasso_model <- cv.glmnet(X, y, alpha = 1)

# Add predictions to model_data
predictions <- predict(lasso_model, newx = X, s = 'lambda.min')
model_data$pred_win_pct <- as.vector(predictions) %>% round(2)

```

```{r}

# (Intercept) -0.088814061
# eFG          0.046288911
# def_eFG     -0.045708354
# assist_pct  -0.001714749
# steal_pct    0.042897558
# tov_pct     -0.036113247
# oreb_pct     0.009297308
# dreb_pct     0.008733526
# block_pct   -0.009524545

```

```{r}

# Calculate results

# get relevant playoff data
playoff_m <- playoff_summary %>%
  filter(season %in% 2014:2022) %>%
  select(all_of(c(
    "season", "off_team", "def_team", "Round", 'series_winner'
  )))

# add win percentages to playoff data
playoff_m <- playoff_m %>%
  inner_join(model_data[, c('season', 'off_team', 'pred_win_pct')], 
             by = c('season', 'off_team')) %>%
  rename(off_win_pct = pred_win_pct) %>%
  inner_join(model_data[, c('season', 'off_team', 'pred_win_pct')], 
             by = c('season', 'def_team' = 'off_team')) %>%
  rename(def_win_pct = pred_win_pct)

# normalize probabilities for each series

playoff_m <- playoff_m %>%
  mutate(p_total = off_win_pct + def_win_pct, 
         off_win_pct = off_win_pct/p_total,
         def_win_pct = def_win_pct/p_total) %>%
  select(-p_total)

# Add probabilities for possible series results
calc_score_prob <- function(p, total_games) {
  if (total_games == 4) {
    return(p ^ 4)
  } else if (total_games == 5) {
    return(p ^ 4 * (1 - p) * 4)
  } else if (total_games == 6) {
    return(p ^ 4 * (1 - p) ^ 2 * 10)
  } else if (total_games == 7) {
    return(p ^ 4 * (1 - p) ^ 3 * 20)
  }
}

playoff_m <- playoff_m %>%
  mutate(
    off_in_4 = calc_score_prob(off_win_pct, 4),
    off_in_5 = calc_score_prob(off_win_pct, 5),
    off_in_6 = calc_score_prob(off_win_pct, 6),
    off_in_7 = calc_score_prob(off_win_pct, 7),
    def_in_4 = calc_score_prob(def_win_pct, 4),
    def_in_5 = calc_score_prob(def_win_pct, 5),
    def_in_6 = calc_score_prob(def_win_pct, 6),
    def_in_7 = calc_score_prob(def_win_pct, 7),
    off_advances = off_in_4 + off_in_5 + off_in_6 + off_in_7,
    def_advances = def_in_4 + def_in_5 + def_in_6 + def_in_7
  ) %>%
  mutate(across(6:17, ~round(.x, 2)))


final_output <- playoff_m %>% select(-c('off_win_pct', 'def_win_pct'))


```

## Model Overview

I used a linear model to predict win percentage of teams for the regular season based on a number of advanced stats. These win percentages are then used as an estimate of each team's winning probability of each game during a playoff series. The probabilities of all possible series results and each team advancing are calculated using the estimates on individual games. 

Some notable insights from the results are:

- Increasing offensive efficiency and decreasing opponents offensive efficiency has a similar impact
- Similarly defensive rebounds and offensive rebound efficiency has similar impact
- No meaningful relationship between assist percentage and number of wins
- Block percentage has a negative association with number of wins. This should be further investigated.

### Strengths of the model

- The contribution of each considered stat to the final winning percentage is easily interpreted helping ease decision making 
- The results can be understood by a wide audience.
- Estimates the probabilities for each game rather than estimating than estimating the results of the series

### Weaknesses of the model

- The assumption that regular season winning percentage is a good estimate of the winning percentage of each game during the playoffs is not verified
- The model uses a small number of advanced stats.
- The model doesn't provide any estimate of the advanced stats against a specific opponent in the playoffs
- No component that measures the team's recent form before the playoffs

### Possible improvements

- Using playoff data from previous years, a separate model can be fit to estimate the winning probabilities in a playoff game from the winning percentages during regular season
- We can work with more stats, find their contributions to the winning percentage and explore how they are related to the other stats
- Another component to the estimate of the win probability can be added using stats from more recent games to represent a contribution of recent form

### 2023-24 playoff results estimation

```{r}

reg_2023 <- team_reg_totals %>%
  select(all_of(c(team_info, model_vars))) %>%
  filter(season == 2023)

X <- as.matrix(reg_2023[, model_vars])

predictions <- predict(lasso_model, newx = X, s = 'lambda.min')
reg_2023$pred_win_pct <- as.vector(predictions) %>% round(2)

# create playoff data for 2023

playoff_2023 <- tibble(
  off_team = c("OKC", "BOS", "DEN", "MIN", "PHI", "LAC", "CLE", "MIL"),
  def_team = c("NOP", "MIA", "LAL", "PHX", "NYK", "DAL", "ORL", "IND")
)

# add win percentages to 2023 playoff data
playoff_2023 <- playoff_2023 %>%
  inner_join(reg_2023[, c('off_team', 'pred_win_pct')], 
             by = 'off_team') %>%
  rename(off_win_pct = pred_win_pct) %>%
  inner_join(reg_2023[, c('off_team', 'pred_win_pct')], 
             by = c('def_team' = 'off_team')) %>%
  rename(def_win_pct = pred_win_pct)

playoff_2023 <- playoff_2023 %>%
  mutate(p_total = off_win_pct + def_win_pct, 
         off_win_pct = off_win_pct/p_total,
         def_win_pct = def_win_pct/p_total) %>%
  select(-p_total)

playoff_2023 <- playoff_2023 %>%
  mutate(
    off_in_4 = calc_score_prob(off_win_pct, 4),
    off_in_5 = calc_score_prob(off_win_pct, 5),
    off_in_6 = calc_score_prob(off_win_pct, 6),
    off_in_7 = calc_score_prob(off_win_pct, 7),
    def_in_4 = calc_score_prob(def_win_pct, 4),
    def_in_5 = calc_score_prob(def_win_pct, 5),
    def_in_6 = calc_score_prob(def_win_pct, 6),
    def_in_7 = calc_score_prob(def_win_pct, 7),
    off_advances = off_in_4 + off_in_5 + off_in_6 + off_in_7,
    def_advances = def_in_4 + def_in_5 + def_in_6 + def_in_7
  ) %>%
  mutate(across(5:14, ~round(.x, 2))) %>%
  select(-c(off_win_pct, def_win_pct))

knitr::kable(playoff_2023)

```

